---
title: "Data Visualizations with ggplot"
author: "H.Holt"
date: "2027-09-26"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This week is all about making plots and graphs in ggplot2. The only datasets you need are in the /data directory.

1)  sttstj_fish_biomass
2)  taxonomic

```{r}
#load libraries
library(tidyverse)

#import data
biomass_fish <- read_csv("data/sttstj_fish_biomass.csv")

taxo <- read_csv("data/taxonomic.csv")
```

Inspect the sttstj_fish_biomass dataset. It is 3 years of biomass (kg/177m2) data for 9 fish species inside the VI national park (prot = 1) and outside the national park (prot = 0).

In this instance, anytime you need to group by and summarize, you can just take the mean of biomass and SE. This would not always be appropriate, but for this case the exercises are for plotting and graphing.

\*\*put all your code in the .Rmd file inside a code chunk for each question. Annotate what you are doing so that you can reference them later.

Load Data:

```{r}
#load libraries
library(tidyverse)

#import data
biomass_fish <- read_csv("data/sttstj_fish_biomass.csv")

taxo <- read_csv("data/taxonomic.csv")
```

Join Data:

```{r}
biomass_species <- left_join(biomass_fish, taxo, by = "SPECIES_CD")
```

1.  Create a line graph of biomass over time for Red hind ("EPI GUTT") inside the national park.

```{r}
#filter by species and protected status, keep needed columns
redhind_biomass_NP <- biomass_species %>% 
  filter(COMNAME.x == "red hind") %>% 
  filter(protected_status == "1") %>%
  select(YEAR, biomass, SE, protected_status, COMNAME.x)

#graph bby
ggplot (data = redhind_biomass_NP, 
  aes(x = YEAR, y = biomass)) + 
  geom_line(color = "red", linetype = 1)
```

2.  Create a point and line graph of biomass over time for Yellowtail Snapper ("OCY CHRY") outside the national park.

```{r}
#use biomass_species and filter accordingly
yts_biomass_ONP <- biomass_species %>% 
  filter(COMNAME.x == "yellowtail snapper") %>% 
  filter(protected_status == "0") %>%
  select(YEAR, biomass, SE, protected_status, COMNAME.x)

#graph that data!
ggplot (data = yts_biomass_ONP, 
  aes(x = YEAR, y = biomass)) + 
  geom_line(color = "#dcc516", linetype = 1) +
  geom_point(color = "#dcc516")
```

3.  Create a bar graph of biomass over time for each species grouped by protected_status. Edit the the axis and legend titles to be more insightful.

    -   biomass units
    -   Scientific name instead of SPECIES_CD
    -   0 = Open, 1 = VI National Park

    ```{r}
    #note: I think it would be better to have two graphs one in the park and one out to compare but here is a graph with both
    ggplot (data = biomass_species, 
      aes(x = interaction(YEAR, protected_status), y = biomass, fill = SCINAME )) +
      geom_col(position = position_dodge()) + 
      scale_y_continuous(expand = c(0,0))

    #problem bc protected status is being treated like a continuous variable, fix by making a factor
    biomass_species$protected_status <- factor(biomass_species$protected_status)

    #now graph, used Gemini to help format 
    ggplot (data = biomass_species, 
        aes(x = interaction(YEAR, protected_status), y = biomass, fill = SCINAME)) +
      geom_col(position = position_dodge()) + 
      scale_y_continuous(expand = c(0,0)) +
      
      # change view of x axis: labeling interaction values
      scale_x_discrete(
    name = "Year and Protected Status", # x axis title

    #use the labels argument to map the interaction (e.g., "2010.0") to a clearer label
    # This requires knowing all unique YEAR.protected_status combinations.
    # The following function dynamically creates them:
    labels = function(x) {
      parts <- strsplit(x, "\\.")
      sapply(parts, function(p) {
        year <- p[1]
        status_code <- p[2]
        status_label <- switch(status_code, 
                               "0" = "Open", 
                               "1" = "VI National Park", 
                               status_code)
        paste(year, "\n", status_label)})}) +
      
      #edit axis and legend labels
      labs(
    y = "Biomass (kg/177m2)", 
    fill = "Scientific Name") +
      theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
    ```

4.  Create a bar graph of trophic-level mean biomass inside and outside VI national park for each species.

```{r}
#need to get mean biomass 
mean_biomass <- biomass_species %>%
  group_by(trophic, protected_status) %>%
  summarise(
    mean_biomass = mean(biomass, na.rm = TRUE),
    .groups = 'drop')

#now graph
ggplot (data = mean_biomass, 
  aes(x = trophic, y = mean_biomass, fill = protected_status)) +
  geom_col(position = position_dodge()) + 
  scale_fill_discrete(labels = c("Open", "VI National Park")) + 
  scale_y_continuous(expand = c(0,0)) + 
  labs(x = "Trophic Level", y = "Mean Biomass", fill = "Management Area")
```

5.  Create a point and line plot of 'grouper' biomass (+- SE) over time; inside and outside park together.

```{r}
#mean grouper biomass, across all species 
grouper_biomass <- biomass_species %>%
          filter(trophic == "grouper") %>% 
  group_by(YEAR, protected_status) %>%
  summarise(
    mean_biomass = mean(biomass, na.rm = TRUE),
    mean_se = mean(SE, na.rm = TRUE),
    .groups = 'drop'
  )

#graph
ggplot(data = grouper_biomass,
         aes(x = YEAR, y = mean_biomass, color = protected_status)) +
  geom_line(linetype = 1) +
  geom_errorbar(aes(ymin = mean_biomass - mean_se,
                    ymax = mean_biomass + mean_se),
                width = 0.2) + 
  scale_color_discrete(labels = c("Open", "VI National Park")) +
  labs(
    title = "Mean Grouper Biomass Over Time Within and Outside of Protected Areas",
    x = "Year",
    y = "Mean Biomass (+/- SE)",
    color = "Management Area"
  ) +
  theme_minimal()
```
