---
title: "Data Visualizations with ggplot"
author: "J.Blondeau"
date: "2027-07-27"
output: html_document
---

```{r setup, echo=FALSE}
knitr::opts_chunk$set(echo = FALSE,
                      warning = FALSE,
                      message = FALSE)
options(width = 750)
```

Load the fish data
```{r}
library(tidyverse)
sttstj_fish <- read_csv("../data/sttstj_fish_biomass.csv")
```

1. Create a line graph of biomass over time for Red hind ("EPI GUTT") inside the national park.

```{r}
sttstj_fish %>%
  filter(SPECIES_CD == "EPI GUTT", protected_status == 1) %>%
  ggplot(aes(x = YEAR, y = biomass, group = SPECIES_CD)) +
  geom_line() +
  labs(title = "Red Hind biomass over time inside VI National Park",
       x = "Year",
       y = "Biomass")
```


2. Create a point and line graph of biomass over time for Yellowtail Snapper ("OCY CHRY") outside the national park.

```{r}
sttstj_fish %>%
  filter(SPECIES_CD == "OCY CHRY", protected_status == 0) %>%
  ggplot(aes(x = YEAR, y = biomass, group = SPECIES_CD)) +
  geom_point() +
  geom_line() +
  labs(title = "Yellowtail Snapper biomass over time outside VI National Park",
       x = "Year",
       y = "Biomass")

```


3. Create a bar graph of biomass over time for all the parrotfish combined grouped by protected_status.  Edit the the axis and legend titles to be more insightful.
    - biomass units
    - 0 = Open, 1 = VI National Park
    
```{r}
sttstj_fish %>%
  filter(trophic == "parrotfish") %>% 
  group_by(YEAR, protected_status) %>%
  summarise(biomass = sum(biomass)) %>%
  ggplot(aes(x = YEAR, y = biomass, fill = factor(protected_status))) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Parrotfish biomass",
       x = "Year",
       y = "Biomass (kg/177m2)") +
  scale_fill_manual(values = c("0" = "red", "1" = "blue"),
                    name = element_blank(),
                    labels = c("Outside", "Inside"))+
  scale_x_continuous(breaks = c(2017, 2019, 2021, 2023)) +
  scale_y_continuous(expand = c(0, 0))
```

    
4. Create a bar graph of trophic-level mean biomass inside and outside VI national park.

```{r}
sttstj_fish %>% 
  group_by(YEAR, trophic, protected_status) %>% 
  summarise(tot_bio = sum(biomass)) %>% 
  group_by(trophic, protected_status) %>% 
  summarise(mean_bio = mean(tot_bio)) %>%
  ggplot(aes(x = trophic, y = mean_bio, fill = factor(protected_status))) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Mean biomass by trophic level",
       x = "Trophic level",
       y = "Mean biomass (kg/177m2)") +
  scale_fill_manual(values = c("0" = "red", "1" = "blue"),
                    name = element_blank(),
                    labels = c("Outside", "Inside")) +
  scale_y_continuous(expand = c(0, 0))

```


5. Create a point and line plot of 'grouper' biomass (+- SE) over time; inside and outside park together.

```{r}
sttstj_fish %>%
  group_by(YEAR, protected_status, trophic) %>%
  summarise(tot_bio = sum(biomass), SE = sum(SE)) %>%
  filter(trophic == "grouper") %>%
  ggplot(aes(x = YEAR, y = tot_bio, group = protected_status, color = factor(protected_status))) +
  geom_errorbar(aes(ymin = tot_bio - SE, ymax = tot_bio + SE), width = 0.1) +
  geom_point() +
  geom_line() +
  theme_minimal() +
  labs(title = "Grouper biomass over time",
       x = "Year",
       y = "Biomass (kg/177m2)") +
  scale_color_manual(values = c("0" = "red", "1" = "blue"),
                     name = element_blank(),
                     labels = c("Outside", "Inside")) +
  scale_x_continuous(breaks = c(2017, 2019, 2021, 2023))
```

